---
title: "The Effect of Beer Taxes on Traffic Fatalities"
author: "Christopher Pellettiere, Alexander Acevedo, Brandon Kim"
date: "June 5, 2026"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6.5,
  fig.height = 3.5
)

# Packages
library(AER)
library(plm)
library(tidyverse)
library(stargazer)
library(lmtest)
library(sandwich)
library(car)
library(kableExtra)
```

# Panel Data Model

## Economic Question and Dataset

This project investigates whether higher beer taxes are associated with lower traffic fatality rates across U.S. states over time. The main economic question is:

**Do higher beer taxes reduce traffic fatalities?**

This question is important because alcohol-related driving fatalities create large social costs, including loss of life, medical costs, property damage, and enforcement costs. If higher beer taxes are associated with fewer traffic fatalities, then alcohol taxation may be a useful public safety policy, not just a revenue-generating policy.

The dataset used in this project is the `Fatalities` dataset from the `AER` package in R. The dataset is a balanced state-level panel that follows 48 U.S. states from 1982 to 1988, for a total of 336 state-year observations. The dependent variable is a traffic fatality rate, constructed as traffic fatalities per 10,000 people. The main explanatory variable of interest is the beer tax. Additional control variables include the minimum legal drinking age, income, miles driven, and population.

Panel data is appropriate for this question because traffic fatality rates likely differ across states for persistent reasons that are difficult to fully observe. For example, states may differ in driving culture, rural road exposure, alcohol consumption norms, weather, law enforcement intensity, public transportation availability, road quality, and baseline attitudes toward drunk driving. These state-specific factors may affect traffic fatalities and may also be correlated with beer taxes or alcohol policies.

Because of this, a pooled OLS model may suffer from omitted variable bias if it ignores unobserved state heterogeneity. A fixed effects model can control for time-invariant state characteristics by comparing each state to itself over time. A random effects model may be more efficient, but it requires the stronger assumption that the unobserved state-specific effects are uncorrelated with the explanatory variables. This project estimates pooled OLS, fixed effects, and random effects models, then uses formal statistical tests to determine the preferred specification.

```{r part-a-panel-structure}
data("Fatalities", package = "AER")

fatalities_raw <- Fatalities

fatalities_pd <- pdata.frame(
  fatalities_raw,
  index = c("state", "year")
)

num_states <- length(unique(fatalities_raw$state))
num_years <- length(unique(fatalities_raw$year))
num_obs <- nrow(fatalities_raw)
balanced_status <- is.pbalanced(fatalities_pd)

panel_structure <- data.frame(
  `Number of States` = num_states,
  `Number of Years` = num_years,
  `Total Observations` = num_obs,
  `Balanced Panel` = balanced_status
)

kable(
  panel_structure,
  caption = "Panel Structure of the Fatalities Dataset"
)
```

## Summary Statistics and Panel Structure Visualizations

```{r variable-construction}
fatalities_pd <- fatalities_pd %>%
  as.data.frame() %>%
  mutate(
    fatal_rate = fatal / pop * 10000,
    log_income = log(income),
    log_miles = log(miles),
    log_pop = log(pop)
  ) %>%
  pdata.frame(index = c("state", "year"))
```

```{r summary-statistics}
summary_stats <- fatalities_pd %>%
  as.data.frame() %>%
  select(fatal_rate, beertax, drinkage, log_income, log_miles, log_pop) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  group_by(Variable) %>%
  summarise(
    Mean = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),
    Min = min(Value, na.rm = TRUE),
    Max = max(Value, na.rm = TRUE),
    .groups = "drop"
  )

summary_stats <- summary_stats %>%
  mutate(
    Variable = recode(
      Variable,
      "beertax" = "Beer Tax",
      "drinkage" = "Minimum Drinking Age",
      "fatal_rate" = "Fatality Rate",
      "log_income" = "Log Income",
      "log_miles" = "Log Miles Driven",
      "log_pop" = "Log Population"
    )
  )

kable(
  summary_stats,
  digits = 3,
  caption = "Summary Statistics for Model Variables"
)
```

Table 2 reports summary statistics for the variables used in the model. The average traffic fatality rate is 2.040 fatalities per 10,000 people, with values ranging from 0.821 to 4.218, showing meaningful variation across state-year observations. Beer taxes also vary substantially, with a mean of 0.513 and a range from 0.043 to 2.721. The minimum legal drinking age ranges from 18 to 21, which reflects policy variation across states and years. The income, miles, and population variables are logged to reduce the influence of large differences across states and to make the controls less skewed.

```{r fatality-rate-over-time}
ggplot(as.data.frame(fatalities_pd), aes(x = year, y = fatal_rate, group = state)) +
  geom_line(alpha = 0.35) +
  labs(
    title = "Traffic Fatality Rates Across States Over Time",
    x = "Year",
    y = "Fatalities per 10,000 People"
  )
```
**Figure 1: Traffic Fatality Rates Across States Over Time.**

Figure 1 shows traffic fatality rates for each state from 1982 to 1988. The lines differ in their average levels, which suggests individual/state heterogeneity in fatality rates. There is also movement over time within states, meaning the dependent variable has both cross-sectional and time-series variation. This supports the use of panel methods and motivates testing whether pooled OLS is sufficient or whether state effects should be included.


```{r beer-tax-over-time}
ggplot(as.data.frame(fatalities_pd), aes(x = year, y = beertax, group = state)) +
  geom_line(alpha = 0.35) +
  labs(
    title = "Beer Taxes Across States Over Time",
    x = "Year",
    y = "Beer Tax"
  )
```
**Figure 2: Beer Taxes Across States Over Time.**  

Figure 2 shows beer taxes for each state over time. Some states consistently have higher beer taxes than others, suggesting cross-state differences in alcohol tax policy. The graph also shows some within-state changes over time, although beer taxes are relatively stable for many states. This is important because the fixed effects model 
identifies the effect of beer taxes using within-state changes over time.


```{r beer-tax-fatality-scatter}
ggplot(as.data.frame(fatalities_pd), aes(x = beertax, y = fatal_rate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Beer Taxes and Traffic Fatality Rates",
    x = "Beer Tax",
    y = "Fatalities per 10,000 People"
  )
```
**Figure 3: Pooled Relationship Between Beer Taxes and Traffic Fatality Rates.** 

Figure 3 shows the pooled relationship between beer taxes and traffic fatality rates. The fitted line slopes upward, meaning that higher beer tax observations tend to be associated with higher fatality rates in the raw pooled data. This does not imply that beer taxes increase fatalities. Instead, it suggests that states with higher fatality risks may also have higher beer taxes or other state-specific characteristics. This motivates the use of fixed effects and random effects models to account for unobserved state heterogeneity.

## Model Specification

Before estimating the pooled, fixed effects, and random effects models, we use the following baseline specification:

$FatalRate_{it} = \beta_0 + \beta_1 BeerTax_{it} + \beta_2 DrinkAge_{it} + \beta_3 \log(Income_{it}) + \beta_4 \log(Miles_{it}) + \beta_5 \log(Population_{it}) + a_i + u_{it}$

where $i$ indexes states and $t$ indexes years.

The dependent variable is $FatalRate_{it}$, which measures traffic fatalities per 10,000 people in a given state and year. This is the outcome we are trying to explain because the project’s main question is whether alcohol taxation is associated with safer driving outcomes.

The main independent variable of interest is $BeerTax_{it}$, the beer tax in each state-year observation. We include this variable because beer taxes may reduce alcohol consumption, especially excessive alcohol consumption, which could reduce drunk driving and lower traffic fatalities. The expected sign on $BeerTax_{it}$ is negative.

The model also includes $DrinkAge_{it}$, the minimum legal drinking age. This is included because legal access to alcohol may directly affect alcohol-related driving behavior, especially among younger drivers. The expected sign is negative because a higher drinking age should reduce access to alcohol for younger drivers and may lower fatality rates.

We include $\log(Income_{it})$ to control for state economic conditions. Income could affect fatality rates in more than one direction. Higher-income states may have safer cars, better roads, and stronger emergency response systems, which could reduce fatalities. However, higher income could also increase vehicle ownership and travel, which could increase accident exposure. Therefore, the expected sign is ambiguous.

We include $\log(Miles_{it})$ to control for driving exposure. States where residents drive more miles have more opportunities for accidents and fatalities to occur. The expected sign is positive.

We include $\log(Population_{it})$ to control for state size and differences between small and large states that may not be fully captured by the fatality rate. Since the dependent variable is already scaled by population, the expected sign is ambiguous.

A simple pooled OLS model may be biased because there are likely unobserved state-specific factors that affect fatality rates and may also be correlated with beer taxes. These potential confounders include driving culture, rural road exposure, road quality, weather, law enforcement intensity, public transportation availability, alcohol consumption norms, and attitudes toward drunk driving. For example, states with higher baseline drunk-driving risk may choose higher beer taxes, which could make the raw relationship between beer taxes and fatalities appear positive even if beer taxes reduce fatalities within a state over time.

The term $a_i$ represents these unobserved state-specific factors that are constant over time. The fixed effects model controls for $a_i$ by comparing each state to itself over time, while the random effects model assumes that $a_i$ is uncorrelated with the regressors. The model-selection tests later in the report will help determine whether pooled OLS, fixed effects, or random effects is preferred.

```{r variable-definition-table}
variable_definitions <- data.frame(
  Variable = c(
    "Fatality Rate",
    "Beer Tax",
    "Minimum Drinking Age",
    "Log Income",
    "Log Miles Driven",
    "Log Population"
  ),
  Role = c(
    "Dependent variable",
    "Main variable of interest",
    "Control",
    "Control",
    "Control",
    "Control"
  ),
  Reason = c(
    "Measures traffic fatalities per 10,000 people",
    "Tests whether alcohol taxation is associated with lower fatality rates",
    "Controls for legal access to alcohol",
    "Controls for state economic conditions",
    "Controls for driving exposure",
    "Controls for state size"
  ),
  `Expected Sign` = c(
    "N/A",
    "Negative",
    "Negative",
    "Ambiguous",
    "Positive",
    "Ambiguous"
  )
)

kable(
  variable_definitions,
  caption = "Model Variables, Roles, and Expected Signs",
  booktabs = TRUE,
  escape = FALSE
) %>%
  kable_styling(
    latex_options = c("hold_position"),
    font_size = 9
  ) %>%
  column_spec(1, width = "1.2in") %>%
  column_spec(2, width = "1.3in") %>%
  column_spec(3, width = "2.5in") %>%
  column_spec(4, width = "1.0in")
```

## Estimation of Panel Data Models

To examine the statistical relationship between beer taxes and traffic fatality rates (as defined above), we estimate three separate panel data specifications: (i) Pooled OLS, (ii) Fixed Effects (FE), and (iii) Random Effects (RE). Each model treats unobserved state-specific heterogeneity differently. Hence, comparing the results across models will help determine whether controlling for unobserved state (time-invariant) characteristics meaningfully changes the estimated effect of beer taxes on traffic fatalities.

### Pooled OLS Model

First, the pooled OLS model treats the panel dataset as a single combined sample and ignores potential unobserved state-specific effects. In other words, it assumes that any such unobserved heterogeneity across states is either negligible or uncorrelated with the explanatory variables (i.e., $\alpha_i = 0$). As such, the intercept and slope coefficients are constant (identical) for all states and time periods, which may not make sense from an economic and statistical viewpoint.

```{r pooled-model-estimation}
pool_model <- plm(
  fatal_rate ~ beertax + drinkage + log_income + log_miles + log_pop,
  data = fatalities_pd,
  model = "pooling"
)
summary(pool_model)
```

```{r pooled-model-table}
pool_results <- broom::tidy(pool_model)

pool_results <- pool_results %>%
  mutate(
    term = recode(
      term,
      "(Intercept)" = "Constant",
      "beertax" = "Beer Tax",
      "drinkage" = "Minimum Drinking Age",
      "log_income" = "Log Income",
      "log_miles" = "Log Miles Driven",
      "log_pop" = "Log Population"
    )
  ) %>%
  select(
    Variable = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t-Statistic` = statistic,
    `P-Value` = p.value
  )

kable(
  pool_results,
  digits = 3,
  caption = "Pooled OLS Model Results"
)
```

Table 4 reports the pooled OLS estimation results for the relationship between beer taxes and traffic fatality rates. Under this specification, the estimated coefficient on $BeerTax$ is positive (0.050), suggesting that higher beer taxes are associated with slightly higher traffic fatality rates in the pooled data. However, the coefficient is not statistically significant ($p = 0.3716$), meaning there is insufficient evidence to conclude that beer taxes have a measurable effect on fatality rates in the pooled OLS model.

Among the control variables, $\log({Income})$ and $\log({Miles})$ are both statistically significant at the 1% level. In particular, higher income is associated with lower fatality rates (coefficient of -1.551), while higher driving exposure is associated with higher fatality rates (coefficient of 1.622), which is consistent with our expectation that states where residents drive more miles face greater accident risk. Neither $DrinkAge$ nor $\log({Population})$ are statistically significant in this specification.

In addition, the pooled model has an $R^2$ of 0.46, indicating that approximately 46% of the variation in traffic fatality rates can be explained by the included regressors. However, the model likely suffers from omitted variable bias because it ignores unobserved state-specific characteristics such as driving culture, road quality, weather conditions, law enforcement intensity, and attitudes toward drunk driving. If these unobserved factors are correlated with beer taxes or other regressors, then the provided estimates may be biased and inconsistent. For this reason, the fixed effects and random effects models are estimated next to account for unobserved heterogeneity across states.

### Fixed Effects Model (FEM)

Second, the fixed effects model algebraically controls for unobserved, time-invariant state characteristics through _demeaning_, which also allows each state to have its own intercept (and also allows $Cov(\alpha_i, X_{it}) \neq 0$). Unlike pooled OLS, the FEM removes the influence of all state-specific factors that do not change over time, such as geography (i.e., rural vs urban density) and persistent differences in road infrastructure. Thus, this model estimates the effect of beer taxes using _within-state_ (not between) changes over time. 

```{r fixed-effects-estimation}
fe_model <- plm(
  fatal_rate ~ beertax + drinkage + log_income + log_miles + log_pop,
  data = fatalities_pd,
  model = "within"
)
summary(fe_model)
```

```{r fixed-effects-table}
fe_results <- broom::tidy(fe_model)

fe_results <- fe_results %>%
  mutate(
    term = recode(
      term,
      "beertax" = "Beer Tax",
      "drinkage" = "Minimum Drinking Age",
      "log_income" = "Log Income",
      "log_miles" = "Log Miles Driven",
      "log_pop" = "Log Population"
    )
  ) %>%
  select(
    Variable = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t-Statistic` = statistic,
    `P-Value` = p.value
  )

kable(
  fe_results,
  digits = 3,
  caption = "Fixed Effects Model Results"
)
```

As displayed above, Table 5 reports the fixed effects estimation results. To reiterate, the FEM controls for unobserved time-invariant state characteristics by comparing each state to itself over time, thereby removing the influence of persistent state-specific factors such as driving culture, geographic conditions, and road infrastructure that may be correlated with $BeerTax$ or other regressors. Critically, this means that the FEM eliminates any omitted variable bias from _time-invariant_ factors, as well as the constant intercept ($\beta_0$) we saw in pooled OLS.

Under the FEM, the estimated coefficient on $BeerTax$ is negative (-0.314), consistent with our hypothesis that higher beer taxes reduce traffic fatality rates. In addition, the coefficient is statistically significant at the 10% level ($p = 0.096$), but not at the 5% level, suggesting moderate evidence that increases in beer taxes are associated with lower fatality rates _within_ states over time. This result differs substantially from the pooled OLS model, where the beer tax coefficient was positive and statistically insignificant. From a statistical standpoint, the change in sign suggests that unobserved state heterogeneity may have biased the earlier pooled OLS estimates.

Further, the coefficient on $DrinkAge$ is negative and statistically significant at the 5% level, indicating that increases in the legal drinking age are associated with lower traffic fatality rates. Likewise, $\log({Population})$'s coefficient is also statistically significant and negative, while that for $\log({Income})$ is positive and statistically significant. Unlike in the pooled OLS model, the coefficient on $\log({Miles})$ is not statistically significant after controlling for state fixed effects.

The FEM has an $R^2$ of 0.165, meaning that approximately 16.5% of the within-state variation in traffic fatality rates is explained by the regressors. Although the within $R^2$ is lower than the pooled OLS $R^2$, the fixed effects estimator is still  more credible when unobserved state-specific effects are correlated with the explanatory variables.

### Random Effects Model (REM)

Finally, the random effects model accounts for unobserved state-specific heterogeneity while assuming that such unobserved state effects are _uncorrelated_ with the explanatory variables (i.e., $Cov(\alpha_i, X_{it}) = 0$), which is often an unrealistic assumption. Unlike the fixed effects model, the REM treats any heterogeneity as part of the error term ($u_{it}$) and thus uses both within-state and between-state variation. To be more specific, the REM assumes each state's intercept differs randomly from the population average, i.e., $\beta_{0i} = \bar{\beta_0} + u_i$ If assumptions hold, the random effects estimators can be more efficient. 

```{r random-effects-estimation}
re_model <- plm(
  fatal_rate ~ beertax + drinkage + log_income + log_miles + log_pop,
  data = fatalities_pd,
  model = "random"
)
summary(re_model)
```

```{r random-effects-table}
re_results <- broom::tidy(re_model)

re_results <- re_results %>%
  mutate(
    term = recode(
      term,
      "(Intercept)" = "Constant",
      "beertax" = "Beer Tax",
      "drinkage" = "Minimum Drinking Age",
      "log_income" = "Log Income",
      "log_miles" = "Log Miles Driven",
      "log_pop" = "Log Population"
    )
  ) %>%
  select(
    Variable = term,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t-Statistic` = statistic,
    `P-Value` = p.value
  )

kable(
  re_results,
  digits = 3,
  caption = "Random Effects Model Results"
)
```

Table 6 reports the random effects estimation results. Under this final specification, the estimated coefficient on $BeerTax$ is positive (0.188) and statistically significant at the 10% level ($p = 0.098$), once again suggesting moderate evidence that higher beer taxes are associated with higher traffic fatality rates. This result differs from the fixed effects model, where the beer tax coefficient was negative (but still significant). Critically, the difference between the fixed and random effects estimates could indicate that the assumption underlying the random effects model, namely, $Cov(\alpha_i, X_{it}) = 0$, may not hold in this application.

Among the control variables, the slope coefficient on $\log{(Miles)}$  is positive and statistically significant at the 5% level, indicating that states with greater driving exposure tend to experience higher fatality rates in the REM. Also, the coefficient on $\log{(Population)}$ is negative and statistically significant, while the coefficient on $\log{(Income)}$ is positive and marginally significant at the 10% level. $DrinkAge$ is not statistically significant in the random effects specification.

Further, the REM has an $R^2$ of 0.075, indicating that the model explains approximately 7.5% of the variation in traffic fatality rates. Because the validity of the random effects estimator depends on the assumption that unobserved state effects are uncorrelated with the regressors, a formal Hausman test will later be used to compare the fixed and random effects specifications and determine the more appropriate model.

### Comparative Table of Panel Results

Below, we summarize our findings thus far in a single comparative table displaying all three models side-by-side, including slope coefficients and qualitative considerations.

\newpage

```{r comparative-regression-table}
# Default (non-robust) standard errors
pool_se <- sqrt(diag(vcov(pool_model)))
fe_se   <- sqrt(diag(vcov(fe_model)))
re_se   <- sqrt(diag(vcov(re_model)))

# Function for significance stars (now based on classical SEs)
starify <- function(coef, pval) {
  stars <- ifelse(pval < 0.01, "***",
           ifelse(pval < 0.05, "**",
           ifelse(pval < 0.10, "*", "")))
  
  paste0(sprintf("%.3f", coef), stars)
}

# Classical (non-robust) p-values
pool_p <- coeftest(pool_model, vcov = vcov(pool_model))[,4]
fe_p   <- coeftest(fe_model,   vcov = vcov(fe_model))[,4]
re_p   <- coeftest(re_model,   vcov = vcov(re_model))[,4]

comparison_table <- data.frame(
  Variable = c(
    "Beer Tax", "",
    "Drinking Age", "",
    "Log Income", "",
    "Log Miles", "",
    "Log Population", "",
    "Constant", "",
    "R-Squared",
    "State Fixed Effects?",
    "Random Effects?"
  ),

  `Pooled OLS` = c(
    starify(coef(pool_model)["beertax"], pool_p["beertax"]),
    paste0("(", sprintf("%.3f", pool_se["beertax"]), ")"),

    starify(coef(pool_model)["drinkage"], pool_p["drinkage"]),
    paste0("(", sprintf("%.3f", pool_se["drinkage"]), ")"),

    starify(coef(pool_model)["log_income"], pool_p["log_income"]),
    paste0("(", sprintf("%.3f", pool_se["log_income"]), ")"),

    starify(coef(pool_model)["log_miles"], pool_p["log_miles"]),
    paste0("(", sprintf("%.3f", pool_se["log_miles"]), ")"),

    starify(coef(pool_model)["log_pop"], pool_p["log_pop"]),
    paste0("(", sprintf("%.3f", pool_se["log_pop"]), ")"),

    starify(coef(pool_model)["(Intercept)"], pool_p["(Intercept)"]),
    paste0("(", sprintf("%.3f", pool_se["(Intercept)"]), ")"),

    sprintf("%.3f", summary(pool_model)$r.squared[1]),
    "No",
    "No"
  ),

  `Fixed Effects` = c(
    starify(coef(fe_model)["beertax"], fe_p["beertax"]),
    paste0("(", sprintf("%.3f", fe_se["beertax"]), ")"),

    starify(coef(fe_model)["drinkage"], fe_p["drinkage"]),
    paste0("(", sprintf("%.3f", fe_se["drinkage"]), ")"),

    starify(coef(fe_model)["log_income"], fe_p["log_income"]),
    paste0("(", sprintf("%.3f", fe_se["log_income"]), ")"),

    starify(coef(fe_model)["log_miles"], fe_p["log_miles"]),
    paste0("(", sprintf("%.3f", fe_se["log_miles"]), ")"),

    starify(coef(fe_model)["log_pop"], fe_p["log_pop"]),
    paste0("(", sprintf("%.3f", fe_se["log_pop"]), ")"),

    "",
    "",

    sprintf("%.3f", summary(fe_model)$r.squared[1]),
    "Yes",
    "No"
  ),

  `Random Effects` = c(
    starify(coef(re_model)["beertax"], re_p["beertax"]),
    paste0("(", sprintf("%.3f", re_se["beertax"]), ")"),

    starify(coef(re_model)["drinkage"], re_p["drinkage"]),
    paste0("(", sprintf("%.3f", re_se["drinkage"]), ")"),

    starify(coef(re_model)["log_income"], re_p["log_income"]),
    paste0("(", sprintf("%.3f", re_se["log_income"]), ")"),

    starify(coef(re_model)["log_miles"], re_p["log_miles"]),
    paste0("(", sprintf("%.3f", re_se["log_miles"]), ")"),

    starify(coef(re_model)["log_pop"], re_p["log_pop"]),
    paste0("(", sprintf("%.3f", re_se["log_pop"]), ")"),

    starify(coef(re_model)["(Intercept)"], re_p["(Intercept)"]),
    paste0("(", sprintf("%.3f", re_se["(Intercept)"]), ")"),

    sprintf("%.3f", summary(re_model)$r.squared[1]),
    "No",
    "Yes"
  )
)

# Add assumptions row
comparison_table <- rbind(
  comparison_table,
  c(
    "Assumptions",
    "No state-specific effects (ignored)",
    "Cov($a_i$, $X_{it}$) unrestricted",
    "Cov($a_i$, $X_{it}$) = 0"
  )
)

kable( comparison_table, caption = "Comparison of Panel Data Models", align = c("l", "c", "c", "c"), escape = FALSE )
```

